Compositional equivalences based on Open pNets

### 2021-08-30 2016-Weak-FH-Bisim/WeakBisim.tex
more on substitutions

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{Note that the fact that substitutions are defined from partial function  implies that each $y_i$ is different.}.
 $\subst{\Post}$ is thus the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\otimes$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortotimes\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define the composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

@@

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on a set of variables (potentially with only symbolic values). 
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$, a set $V$ of variables, and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$ is the set of holes involved in the transition; $s, s'\in\mathcal{S}$ are states of the automaton; and $\beta_j$
        is a transition of the hole $j$, with $\Sortop(\beta_j)\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
%        We call \emph{source variables} of an OT the union of all variables
%        in the different terms
%        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the  variables in $V\cup\vars(\alpha)\cup
			{\displaystyle \bigcup_{j\in J'}\vars(beta_j)}$. \Post\ is a set of 
	assignments (i.e., indexed substitutions) that are effective after the open transition, they are
        represented as a substitution of the form $({x_k\gets e_k})^{k\in K}$ 
	where $\forall k.\, x_k\in V$, and $e_k$ are expressions
        over the variables variables in $V\cup\vars(\alpha)\cup
			{\displaystyle \bigcup_{j\in J'}\vars(beta_j)}$. The assignments are always applied simultaneously because the variables in $V$ can be in both sides. Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

@@

For all predicate $\Pred$, for all substitution $\Post$ such that $V\cap \dom(\Post)=\emptyset$. We have: 
		 \begin{mathpar}
    \openrule
         {
           \set{\beta}, \Pred\,',\Post\,'}
          {t \OTarrow {\alpha} t'}\in\mathcal{T}
\implies
    \openrule
         {
           \set{\beta}\subst{\Post}, \Pred\,'\subst{\Post}\land\Pred,\Post\shortotimes\Post\,'}
         {\raisemath{-2pt}{t \OTarrow {\alpha\subst{\Post}} {t'}}{}}
 \in\mathcal{T}
\end{mathpar}
\end{definition}

@@

% \TODO{do we want formulas for this?}
 \end{definition}
Classically, $\Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}$
applies in parallel the  
substitutions $\Post_{OT}$ and $\Post_{OT_x}$ (parallelism is crucial
inside each $\Post$ set but not between  $\Post_{OT}$ and
$\Post_{OT_x}$ that are independent), applying the assignments of the involved rules.
We can prove that such a bisimulation is an equivalence relation.

@@

\end{definition}
Rule $\WTUn$ states that it is always possible to do a non-observable transition, where the state is unchanged and the holes perform no action. Rule~$\WTDeux$ states that each open transition can be considered as a weak open transition. The last rule is the most interesting:  Rule~$\WTTrois$ allows any number of $\tau$ transitions before or after any weak open transition. This rules carefully composes predicates, effects, and actions of the holes, indeed in the rule, predicate $\Pred_2$ manipulates variables of $s_1$ that result from the first weak open transition. Their values thus depend on the initial state but also on the effect (as a substitution $\Post_1$) of the first weak open transition. In the same manner, $\Pred_3$ must be applied the joint substitution $\Post_2\shortotimes\Post_1$. Similarly, effects on variables must be applied to obtain the global effect of the composed weak open transition, it must also be applied to observable actions of the holes, and to the global action of the weak open transition.

@@

Now let us give some details about the construction of the weak automaton of the implementation pNet, obtained by application of the weak rules as explained above. We concentrate on weak open transitions $WI_3$ and $WI_4$. Let us denote as $post_n$ the effect (substitution) of the strong open transitions $SI_n$ from Figure \ref{SimpleProtCounter:ImplOA}:
\smallskip

@@

Then the effect of one single $100 \xrightarrow{OT_4} 210 \xrightarrow{OT_5} 220 \xrightarrow{OT_6} 100$ loop is\footnote{when showing  the result of $Posts$ composition, we will omit the identity substitutions introduced by the $\shortotimes$ definition in page \pageref{def:substitutions}}:
\[post_{456} = post_6 \shortotimes\ post_5 \shortotimes\ post_4
= (\texttt{s\_ec}\gets \texttt{s\_ec}+1)\]

@@

By construction, four substitutions $\subst{~}$ only have an effect on the  
variables of the open automaton they belong to, they also produce terms containing only 
variables of the open automaton they belong to. Finally, because of the domain of the 
substitutions of the predicates, we have:\\
       	$\Pred_{{px}}\subst{\Post_{OT}\uplus\Post_{OT_{px}}}\land 
       	\Pred_{{pxy}}\subst{\Post_{OT_{px}}\uplus
       	\Post_{OT_{pxy}}}\Leftrightarrow\\ % we changed =
       	\Pred_{{px}}\subst{\Post_{OT}\uplus\Post_{OT_{px}}\uplus\Post_{OT_{pxy}}}\land
       	\Pred_{{pxy}}\subst{\Post_{OT}\uplus\Post_{OT_{px}}\uplus\Post_{OT_{pxy}}}\\
       	\implies\Pred\,'_{{pxy}}\subst{\Post_{OT}\uplus
       	\Post_{OT_{px}}\uplus\Post_{OT_{pxy}}}
       	\Leftrightarrow\\ % we changed = 
       	 \Pred\,'_{{pxy}}\subst{\Post_{OT}\uplus\Post_{OT_{pxy}}} $\\

@@

\begin{mathpar}
  \mprset {vskip=.5ex}
\inferrule {
\openrule
         {
           \gamma_{j a}^{j\in J_{a}}, \Pred_{OT_x},\Post_{OT_a}}
         {t \OTWeakarrow {\tau} u_a}
\qquad
\openrule
         {
           \gamma_{j b}^{j\in J_{b}}, \Pred_{OT_{ y}},\Post_{OT_{b}}}
         {u_a \OTWeakarrow {\alpha_2} v_{b}}
\qquad
\openrule
         {
           \gamma_{jc}^{j\in J_{c}}, \Pred_{OT_{c}}, \Post_{OT_{c}}}
         {v_{b} \OTWeakarrow {\tau} w_{c}}
\\ 
\set{\gamma'}=\gamma_{j a}^{j\in J_{a}} \dotcup \gamma_{jb}^{j\in J_{b}} \subst{\Post_{OT_a}} \dotcup\gamma_{jc}^{j\in J_{c}} \subst{\Post_{OT_{ b}}\shortotimes\Post_{OT_a}}\\
\Pred=\Pred_{OT_a}\land\Pred_{OT_{b}}\subst{\Post_{OT_a}}\land \Pred_{OT_{c}}\subst{\Post_{OT_{b}}\shortotimes\Post_{OT_a}}\\
\Post= \Post_{OT_{c}} \shortotimes \Post_{OT_{b}}\shortotimes\Post_{OT_{a}}\quad \alpha''=\alpha_{2}\subst{\Post_{OT_a}}\\
}
{
\openrule
         {
           \set{\gamma},
		\Pred, \Post}  
         {t \OTWeakarrow {\alpha''} w_{c}} 
} 
 \end{mathpar}
It remains to be proven that the following statement holds:\\
$\Pred_{s,t} \land \Pred\implies
 \displaystyle{\bigvee_{x\in X}} 
  ( \forall j\in J. \gamma_j'=\gamma_{j}  \land \Pred
     \land \alpha \!=\! \alpha'' \land  
     \Pred_{s',x}\subst{\Post_{OT}\uplus\Post\,})$\\
We have:\\
$\Pred_{OT}= \Pred_1 \wedge \Pred_2\subst{\Post_1} \wedge \Pred_3\subst{\Post_2\shortotimes\Post_1}$\\
$\Post_{OT}=\Post_3\shortotimes \Post_2\shortotimes\Post_1$\\
Moreover, we have the following  statement:\\
$\Pred_{s,t} \land \Pred_1\implies
 \displaystyle{\bigvee_{a\in A}}
 \! \! \left( \forall j\in J_a. \vis{\beta_{1j}}=\gamma_{ja}  \land \Pred_{OT_a}
      \land  
     \Pred_{s_{11},a}\subst{\Post_1\uplus\Post_{OT_a}}\right)$\\
With the conjunction of the predicate $\Pred_2\subst{\Post_1}$ on both sides of the implication, we get:
\begin{multline*}
(\Pred_{s,t} \land \Pred_1)\wedge\Pred_2\subst{\Post_1}\implies \\
 \bigvee_{a\in A}
  \!\! ( \forall j\in J_a. \vis{\beta_{1j}}\!=\!\gamma_{ja}\!  \land\! \Pred_{OT_a}\! \land\!
     \Pred_{s_{11},a}\subst{\Post_1\uplus\Post_{OT_a}}) \wedge \Pred_2\subst{\Post_1\uplus\Post_{OT_a}}  \hspace{-2ex}
\end{multline*}
Note that  on the right side of the implication we added  the substitution  of $\Post_{OT_x}$ without affecting the validity of the statement, because  the domain  of the substitution $\Post_{OT_x}$ is disjoint from the others. Hence a little rewriting gives:
\begin{multline*}
(\Pred_{s,t} \land \Pred_1)\wedge\Pred_2\subst{\Post_1}\implies\\
\bigvee_{a\in A}
\!\!   \left(\forall j\in J_a. \vis{\beta_{1j}}\!=\!\gamma_{ja}\land \Pred_{OT_a}
     \!\land(\Pred_{s_{11},a}\wedge\Pred_2)\subst{\Post_1\uplus\Post_{OT_a}} \right)
\end{multline*}

@@

Again note that because of the domain  of the substitution is independent from some predicates and expressions, we removed $\Post_1$ and  we  added the term $\Post_{OT_{b}}\shortotimes\,\Post_{OT_a}$ in the substitution of the right side of the implication.\\

>>>>

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{Note that the fact that substitutions are defined from partial function  implies that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following.
 $\subst{\Post}$ is thus the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\otimes$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortotimes\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

@@

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on a set of variables (potentially with only symbolic values). 
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$, a set $V$ of variables, and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$ is the set of holes involved in the transition; $s, s'\in\mathcal{S}$ are states of the automaton; and $\beta_j$
        is a transition of the hole $j$, with $\Sortop(\beta_j)\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
%        We call \emph{source variables} of an OT the union of all variables
%        in the different terms
%        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the  variables in $V\cup\vars(\alpha)\cup
			{\displaystyle \bigcup_{j\in J'}\vars(beta_j)}$. \Post\ is a set of 
	assignments that are effective after the open transition, they are
        represented as a substitution function: $({x_k\gets e_k})^{k\in K}$ 
	where $\forall k.\, x_k\in V$, and $e_k$ are expressions
        over the variables variables in $V\cup\vars(\alpha)\cup
			{\displaystyle \bigcup_{j\in J'}\vars(beta_j)}$. The assignments are always applied simultaneously because the variables in $V$ can be in both sides ($x_k$s are distinct). Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

@@

For all predicate $\Pred$, for all partial functions $\Post$ such that $V\cap \dom(\Post)=\emptyset$. We have: 
		 \begin{mathpar}
    \openrule
         {
           \set{\beta}, \Pred\,',\Post\,'}
          {t \OTarrow {\alpha} t'}\in\mathcal{T}
\implies
    \openrule
         {
           \set{\beta}\subst{\Post}, \Pred\,'\subst{\Post}\land\Pred,\Post\shortotimes\Post\,'}
         {\raisemath{-2pt}{t \OTarrow {\alpha\subst{\Post}} {t'}}{}}
 \in\mathcal{T}
\end{mathpar}
\end{definition}

@@

% \TODO{do we want formulas for this?}
 \end{definition}
Classically, $\Pred_{s',t_x}\subst{\Post_{OT}\uplus\Post_{OT_x}}$
applies in parallel the  
substitution defined by the partial functions $\Post_{OT}$ and $\Post_{OT_x}$ (parallelism is crucial
inside each $\Post$ set but not between  $\Post_{OT}$ and
$\Post_{OT_x}$ that are independent), applying the assignments of the involved rules.
We can prove that such a bisimulation is an equivalence relation.

@@

\end{definition}
Rule $\WTUn$ states that it is always possible to do a non-observable transition, where the state is unchanged and the holes perform no action. Rule~$\WTDeux$ states that each open transition can be considered as a weak open transition. The last rule is the most interesting:  Rule~$\WTTrois$ allows any number of $\tau$ transitions before or after any weak open transition. This rules carefully composes predicates, effects, and actions of the holes, indeed in the rule, predicate $\Pred_2$ manipulates variables of $s_1$ that result from the first weak open transition. Their values thus depend on the initial state but also on the effect (as a substitution function $\Post_1$) of the first weak open transition. In the same manner, $\Pred_3$ must be applied the substitution defined by the composition $\Post_2\shortotimes\Post_1$. Similarly, effects on variables must be applied to obtain the global effect of the composed weak open transition, it must also be applied to observable actions of the holes, and to the global action of the weak open transition.

@@

Now let us give some details about the construction of the weak automaton of the implementation pNet, obtained by application of the weak rules as explained above. We concentrate on weak open transitions $WI_3$ and $WI_4$. Let us denote as $post_n$ the effect (as a substitution function) of the strong open transitions $SI_n$ from Figure \ref{SimpleProtCounter:ImplOA}:
\smallskip

@@

Then the effect of one single $100 \xrightarrow{OT_4} 210 \xrightarrow{OT_5} 220 \xrightarrow{OT_6} 100$ loop is\footnote{when showing  the result of $Posts$ composition, we will omit the identity substitution functions introduced by the $\shortotimes$ definition in page \pageref{def:substitutions}}:
\[post_{456} = post_6 \shortotimes\ post_5 \shortotimes\ post_4
= (\texttt{s\_ec}\gets \texttt{s\_ec}+1)\]

@@

By construction, four substitution functions $\subst{~}$ only have an effect on the  
variables of the open automaton they belong to, they also produce terms containing only 
variables of the open automaton they belong to. Finally, because of the domain of the 
substitution functions of the predicates, we have:\\
       	$\Pred_{{px}}\subst{\Post_{OT}\uplus\Post_{OT_{px}}}\land 
       	\Pred_{{pxy}}\subst{\Post_{OT_{px}}\uplus
       	\Post_{OT_{pxy}}}\Leftrightarrow\\ % we changed =
       	\Pred_{{px}}\subst{\Post_{OT}\uplus\Post_{OT_{px}}\uplus\Post_{OT_{pxy}}}\land
       	\Pred_{{pxy}}\subst{\Post_{OT}\uplus\Post_{OT_{px}}\uplus\Post_{OT_{pxy}}}\\
       	\implies\Pred\,'_{{pxy}}\subst{\Post_{OT}\uplus
       	\Post_{OT_{px}}\uplus\Post_{OT_{pxy}}}
       	\Leftrightarrow\\ % we changed = 
       	 \Pred\,'_{{pxy}}\subst{\Post_{OT}\uplus\Post_{OT_{pxy}}} $\\

@@

\begin{mathpar}
  \mprset {vskip=.5ex}
\inferrule {
\openrule
         {
           \gamma_{j a}^{j\in J_{a}}, \Pred_{OT_x},\Post_{OT_a}}
         {t \OTWeakarrow {\tau} u_a}
\qquad
\openrule
         {
           \gamma_{j b}^{j\in J_{b}}, \Pred_{OT_{ y}},\Post_{OT_{b}}}
         {u_a \OTWeakarrow {\alpha_2} v_{b}}
\qquad
\openrule
         {
           \gamma_{jc}^{j\in J_{c}}, \Pred_{OT_{c}}, \Post_{OT_{c}}}
         {v_{b} \OTWeakarrow {\tau} w_{c}}
\\ 
\set{\gamma'}=\gamma_{j a}^{j\in J_{a}} \dotcup \gamma_{jb}^{j\in J_{b}} \subst{\Post_{OT_a}} \dotcup\gamma_{jc}^{j\in J_{c}} \subst{\Post_{OT_{ b}}\shortotimes\Post_{OT_a}}\\
\Pred=\Pred_{OT_a}\land\Pred_{OT_{b}}\subst{\Post_{OT_a}}\land \Pred_{OT_{c}}\subst{\Post_{OT_{b}}\shortotimes\Post_{OT_a}}\\
\Post= \Post_{OT_{c}} \shortotimes \Post_{OT_{b}}\shortotimes\Post_{OT_{a}}\quad \alpha''=\alpha_{2}\subst{\Post_{OT_a}}\\
}
{
\openrule
         {
           \set{\gamma},
		\Pred, \Post}  
         {t \OTWeakarrow {\alpha''} w_{c}} 
} 
 \end{mathpar}
It remains to be proven that the following statement holds:\\
$\Pred_{s,t} \land \Pred\implies
 \displaystyle{\bigvee_{x\in X}} 
  ( \forall j\in J. \gamma_j'=\gamma_{j}  \land \Pred
     \land \alpha \!=\! \alpha'' \land  
     \Pred_{s',x}\subst{\Post_{OT}\uplus\Post\,})$\\
We have:\\
$\Pred_{OT}= \Pred_1 \wedge \Pred_2\subst{\Post_1} \wedge \Pred_3\subst{\Post_2\shortotimes\Post_1}$\\
$\Post_{OT}=\Post_3\shortotimes \Post_2\shortotimes\Post_1$\\
Moreover, we have the following  statement:\\
$\Pred_{s,t} \land \Pred_1\implies
 \displaystyle{\bigvee_{a\in A}}
 \! \! \left( \forall j\in J_a. \vis{\beta_{1j}}=\gamma_{ja}  \land \Pred_{OT_a}
      \land  
     \Pred_{s_{11},a}\subst{\Post_1\uplus\Post_{OT_a}}\right)$\\
With the conjunction of the predicate $\Pred_2\subst{\Post_1}$ on both sides of the implication, we get:
\begin{multline*}
(\Pred_{s,t} \land \Pred_1)\wedge\Pred_2\subst{\Post_1}\implies \\
 \bigvee_{a\in A}
  \!\! ( \forall j\in J_a. \vis{\beta_{1j}}\!=\!\gamma_{ja}\!  \land\! \Pred_{OT_a}\! \land\!
     \Pred_{s_{11},a}\subst{\Post_1\uplus\Post_{OT_a}}) \wedge \Pred_2\subst{\Post_1\uplus\Post_{OT_a}}  \hspace{-2ex}
\end{multline*}
Note that  on the right side of the implication we added  the substitution  of $\Post_{OT_x}$ without affecting the validity of the statement, because  the domain  of the substitution function $\Post_{OT_x}$ is disjoint from the others. Hence a little rewriting gives:
\begin{multline*}
(\Pred_{s,t} \land \Pred_1)\wedge\Pred_2\subst{\Post_1}\implies\\
\bigvee_{a\in A}
\!\!   \left(\forall j\in J_a. \vis{\beta_{1j}}\!=\!\gamma_{ja}\land \Pred_{OT_a}
     \!\land(\Pred_{s_{11},a}\wedge\Pred_2)\subst{\Post_1\uplus\Post_{OT_a}} \right)
\end{multline*}

@@

Again note that because of the domain  of the substitution function is independent from some predicates and expressions, we removed $\Post_1$ and  we  added the term $\Post_{OT_{b}}\shortotimes\,\Post_{OT_a}$ in the substitution of the right side of the implication.\\
